Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-15 latex/main.tex

@@

%%% The summary (abstract) should consist of a single paragraph of 150 words or fewer.
%%% This is the SMBE abstract. We can rewrite or work off of it, but it needs to be shorter.
Widely used phylogenetic methods to study SARS-CoV-2 evolution assume that genetic recombination is negligible and
viral relationships can be represented by a single tree.
However, important SARS-CoV-2 recombinant lineages have emerged.
A more suitable approach is to use ancestral recombination graphs (ARGs),
which jointly model recombination, mutation, and genetic inheritance in a cohesive framework.
Recent computational advances have made ARG inference possible for datasets of millions of genome sequences.
Building on these new advances, we have developed sc2ts, a scalable method to simultaneously detect recombinants and
reconstruct ARGs from heterochronous SARS-CoV-2 genome sequences.
Using sc2ts, we inferred an ARG of 2.48 million genomes sampled from January, 2020 to February, 2023.
We find that this pandemic-scale ARG unifies known features of SARS-CoV-2 evolution —
the backbone phylogeny, mutational spectra, and major recognized recombinants.
This ARG enables us to explore the evolutionary origins of known and novel recombinants.
Furthermore, the ARG is compactly stored in the tree sequence format using the tskit library and
can be analyzed efficiently and interactively using standard Python tools on a typical laptop.
This combination of capabilities makes sc2ts a powerful tool
for studying the recombination and mutation processes driving SARS-CoV-2 evolution.

>>>>

%%% The summary (abstract) should consist of a single paragraph of 150 words or
%%% fewer.

@@

% NOTE: currently 155 words

Millions of SARS-CoV-2 genome sequences were collected in
response to the COVID-19 pandemic, forming a dataset of unprecedented
depth and richness.
Estimated genealogies are fundamental to
gleaning information from this ocean of data, 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic sequence data is that recombination is negligible and 
the genealogy is a tree.
Recombinant lineages have risen to global prevalance in SARS-CoV-2,
and therefore such single-tree representations are necessarily
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies for SARS-CoV-2
that automaticaly detects and incorporates recombination, 
and apply it to 2.48 million genomes.
We show that the inferred genealogy captures a broad range of 
known evolutionary features and provides novel 
insights into the relationships between previously identified
recombinant lineages.
Together, the inference method, inferred genealogy and 
suite of open-source analysis tools, 
represent a powerful resource for future research.
